### 2020-07-04 chapters/problem-statement.tex
Small corrections

With this chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.

\section{The Problem}
Many real-time stream monitoring systems are static once deployed in a production environment. These systems are configured \textit{a priori} and make assumptions about future streaming data based on the observable past events. The problem is that if these assumptions no longer hold in the future, the stream monitoring systems' performance decays. Thus, the problem at hand is to determine when to reconfigure the system based on an analysis of the drifts in the stream of data. For example, consider a Machine Learning (ML) model that monitors a data stream of credit card transactions and detects fraud. The ML model trains on a reference period and learns from this training period, \textit{i.e.}, makes assumptions based on the analyzed data. In streaming, throughout time, the data patterns will change and the assumptions made will no longer hold (\textit{e.g.}, buying patterns change on holidays), which leads to a decay in model performance. Our objective is to alert data pattern shifts. Our motivation is doing it so that system administrators can reconfigure the system before the system's performance decays. In this example, system reconfiguration is tantamount to model retraining.

\section{State of the Art Issues}
Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time. In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods. However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the desired characteristics, namely:

@@

\section{Proposal}
%2 phases, names and purpose
We devise a two-phased and two-windowed method: the first phase is a batch analysis that makes use of a reference window; and the second phase is a streaming analysis that makes use of a target sliding window. The batch analysis main goal is to compute the reference aggregation for the reference window. The stream analysis main goals are to incrementally maintain the same type of aggregation but over the streaming data and alert when the difference between the reference aggregation and the target sliding one is high.

%aggregations to represent distributions
We choose to use an histogram aggregation with constant memory usage to encode the distribution of the observed values. Furthermore, we devise a sliding window approximate histogram aggregation that is maintainable in constant time in a sliding window fashion, through the use of Exponential Moving Averages (EMAs). Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for ever growing volumes of data.

Given a user-defined reference period and a stream of data to process, we monitor each event's fields and alert if they are considered divergent relative to their reference aggregation. We provide a measure of how divergent each field is. Based on this information, it is our belief that a system administrator would have an easier time determining when to reconfigure the system.

\section{Assumptions}
Our feature monitoring method assumes only that the monitored features are numerical or categorical and that the events arrive in an orderly fashion during streaming. If the latter is not true, our method will still work but results may be imprecise, as our sliding window histogram aggregation relies on Exponential Moving Averages (EMAs), which give different weights to each event according to its age.

@@

\begin{enumerate}[leftmargin=1.75cm, label=\textbf{(RQ\arabic*)}]
    \item What aggregation can we use to properly encode data distributions?
    \item How can we maintain such an aggregation in a sliding window fashion in real-time with constant memory usage?
    \item How can we measure divergence between the reference and sliding window aggregations?
    \item Based on a divergence measure, when should we raise an alert?
\end{enumerate}

>>>>

With this chapter, we further discuss the problem we aim to solve with this thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.

\section{The Problem} \label{sec:theproblem}
Many real-time stream monitoring systems are static once deployed in a production environment. These systems are configured \textit{a priori} and make assumptions about future streaming data based on the observable past events. The problem is that if these assumptions no longer hold in the future, the stream monitoring systems' performance decays. Thus, the problem at hand is to determine when to reconfigure the system based on an analysis of the drifts in the stream of data. For example, consider a Machine Learning (ML) model that monitors a data stream of credit card transactions and detects fraud. The ML model trains on a reference period and learns from this training period, \textit{i.e.}, makes assumptions based on the analyzed data. In streaming, throughout time, the data patterns will change and the assumptions made will no longer hold (\textit{e.g.}, buying patterns change on holidays), which leads to a decay in model performance. Our objective is to alert data pattern shifts. Our motivation is doing it so that system administrators can reconfigure the system before the system's performance decays. In this example, system reconfiguration is tantamount to model retraining.

\section{State of the Art Issues} \label{sec:sotaissues}
Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time. In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods. However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the desired characteristics, namely:

@@

\section{Proposal} \label{sec:proposal}
%2 phases, names and purpose
We devise a two-phased and two-windowed method: the first phase is a batch analysis that makes use of a reference window; and the second phase is a streaming analysis that makes use of a target sliding window. The batch analysis main goal is to compute the reference aggregation for the reference window. The stream analysis main goals are to incrementally maintain the same type of aggregation but over the streaming data and alert when the difference between the reference aggregation and the target sliding one is high.

%aggregations to represent distributions
We choose to use a histogram aggregation with constant memory usage to encode the distribution of the observed values. Furthermore, we devise a sliding window approximate histogram aggregation that is maintainable in constant time in a sliding window fashion, through the use of Exponential Moving Averages (EMAs). Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for ever-growing volumes of data.

Given a user-defined reference period and a stream of data to process, we monitor each event's fields and alert if they are considered divergent relative to their reference aggregation. We provide a measure of how divergent each field is. Based on this information, we believe that a system administrator would have an easier time determining when to reconfigure the system.

\section{Assumptions} \label{sec:assumptions}
Our feature monitoring method assumes only that the monitored features are numerical or categorical and that the events arrive in an orderly fashion during streaming. If the latter is not true, our method will still work but results may be imprecise, as our sliding window histogram aggregation relies on Exponential Moving Averages (EMAs), which give different weights to each event according to its age.

@@

\begin{enumerate}[leftmargin=1.6cm, label=\textbf{(RQ\arabic*)}]
    \item \textbf{What aggregation can we use to properly encode data distributions?}

    We need an aggregation to reduce the entire window contents to a lighter representation and we need one that encodes the distribution of feature values correctly so we can measure the difference between the reference aggregation and the target sliding window aggregation accurately.

    \item \textbf{How can we maintain such an aggregation in a sliding window fashion in real-time with constant memory usage?}

    We need an aggregation that shows constant space complexity so that we can scale the size of our windows as we wish without increasing memory usage. Additionally, we need to find a way of incrementally maintaining this aggregation in real-time.

    \item \textbf{How can we measure divergence between the reference and sliding window aggregations?}

    After we aggregate the contents of the reference and target sliding windows, we need a function that takes both aggregation states and returns a measure of divergence.

    \item \textbf{Based on a divergence measure, when should we raise an alert?}

    Hard-coding a threshold value for this divergence measure is cumbersome, impractical and error-prone. In other words, we need to find a way of automatically defining the divergence threshold that we use to raise alerts.

\end{enumerate}

\section{Summary}
In Section \ref{sec:theproblem}, we formally defined the problem we aim to solve with this thesis: alert streaming data pattern shifts in regards to a static reference period with the intent of helping system administrators reconfigure their systems before their performance declines. Furthermore, we make explicit we work under time and space constraints, inherent to the streaming paradigm.

In Section \ref{sec:sotaissues}, we summarize the main identified issues regarding state of the art solutions and in Section \ref{sec:proposal} we detail how we propose to implement our solution along with the assumptions made (Section \ref{sec:assumptions}).

In Section \ref{sec:hypothesis}, we defined our hypothesis or main question to address as to whether sliding window aggregations and outlier detection methods combined could be used to detect data pattern shifts in data streaming scenarios in real-time and with a low memory footprint. To better discuss our hypothesis, we pose a series of research questions in Section \ref{sec:rqs}.

In Chapter \ref{chap:my-work}, we discuss initial approaches we thought of and reasons to why we discarded them. Finally, we present and detail each phase and component of our stream monitoring method.

### 2020-07-04 chapters/resumo.tex
Small corrections

A deteção de mudanças de padrões nos dados tem sido investigada nas mais diversas áreas e aplicações e diz respeito ao processo de encontrar padrões nos dados que não correspondam ao esperado ou usual. A deteção atempada e correta destas mudanças de padrões permite que medidas imediatas sejam tomadas, prevenindo perdas de rendimento e mantendo a reputação da empresa entre os seus clientes. Como tal e tendo isto em conta, muitas empresas constroem sistemas de deteção de anomalias. Desta maneira, administradores do sistema são capazes de proactivamente ajustar e prevenir potenciais falhas nos seus serviços.

Por exemplo, o sistema de deteção de anomalias da Yahoo serve o propósito de monitorizar e alertar \textit{streams} de eventos temporais em múltiplos casos de uso da Yahoo. Na Microsoft, o seu sistema interno de deteção de anomalias monitoriza milhões de métricas provenientes de diferentes serviçoes internos, permitindo que os engenheiros os resolvam em tempo real antes que estes problemas adquiram proporções maiores.

Muitos sistemas de monitorização em tempo real são estáticos quando implementados num ambiente de produção. Os engenheiros responsáveis por tais sistemas configuram-nos assumindo que os dados que futuramente fluirão pelo sistema se assemelham aos dados observados até ao momento. Eles fazem-no conscientemente e o melhor que sabem com as ferramentas de que dispoem, mantendo sempre em mente que no futuro possivelmente terão de reconfigurar o sistema. Assim, apesar da configuração inicial do sistema ser a que melhor se adequa, ao longo do tempo, devido a desvios nos padrões de dados, o desempenho do sistema estático incialmente implementado decai gradualmente.

Desta forma, pretendemos avaliar o uso de métodos de análise de \textit{streams} eficientes em termos de memória --- \textit{i.e} métodos que usam pouca memóra e calculam os resultados esperados em tempo real --- com o intuito de detetar desvios nos padrões de dados na presença de grandes volumes de dados com padrões sazonais que fluem pelo sistema a grande velocidade.

\vspace*{10mm}\noindent
\textbf{Keywords}: tempo real, streaming, leves, monitorização, volume, velocidade, variedade, sazonalidade

>>>>

\vspace*{10mm}\noindent
\textbf{Keywords}:

### 2020-07-04 chapters/sota.tex
Small corrections

The third and final axis is the \textit{outlier type} the method aims to report. The authors identify three distinct outlier types: point outliers, subsequence outliers and time-series outliers. A point outlier corresponds to a single data point in time that deviates from the expected value. These point outliers can be uni or multivariate, depending on the nature of the method. A univariate point outlier is a single time-dependent variable that stands out from the rest of the time-series. On the other hand, a multivariate point outlier takes into account multiple time-dependent variables in comparison to the remaining time-series. Subsequence outliers refer to a sequence of points in time that together show anomalous behavior, but may not be individually considered outliers. Similarly to point outliers, a subsequence outlier may take into account a single time-dependent variable or multiple ones. The authors classify them as univariate subsequence outliers and multivariate subsequence outliers, respectively. Lastly, time-series outliers refer to entire time-series being outliers themselves. This type of outliers can only be detected if the input data is multivariate, so as to be able to compare multiple time-series.

@@

Track changes is off
Everyone
Track changes for everyone
You
Track changes for You
Guests
Track changes for guests
pedro.cardoso: How is this error calculated?
Jun 18, 2020 2:52 PM
You: Same as above
Jun 20, 2020 5:09 PM • Edit • Delete
Hit Enter to reply
pedro.cardoso: This 3rd option is under-explained compared to the other 2. Please elaborate
Jun 18, 2020 2:53 PM
You: Not sure myself, that is why is so poorly written. Im going to revisit this
Jun 19, 2020 4:42 PM • Edit • Delete
Hit Enter to reply
pedro.cardoso: Thought: This reads like leftover from the original work. Is there a way to better relate your investigation into SWAGs into the methodology you ended up taking?
Jun 30, 2020 10:19 AM
pedro.cardoso: Please revisit this entire section. Particularly the applicability sections of each algorithm in the context of your current approach.
Jun 30, 2020 10:20 AM
Hit Enter to reply
pedro.cardoso: Remove these TODOs. If you want keep a note of them elsewhere, but not as text render being rendered.
Jul 3, 2020 5:28 PM
pedro.cardoso: You will forget to do it with other tasks still needed to be done.
Jul 3, 2020 5:29 PM
Hit Enter to reply
pedro.cardoso: "because they evolved their reference period"
Jul 3, 2020 5:30 PM
Hit Enter to reply
pedro.cardoso: Where is Model Monitoring?
Jul 3, 2020 5:31 PM
Hit Enter to reply
Current file
Overview
16

@@

\subsubsection*{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable to our goals. In fact, we need a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, leaving the user confused on why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

@@

Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series to find out the most unusual ones. Each subsequence is compared to the others using a distance metric, for instance, the Euclidean distance. The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in quadratic time complexity. These methods find the most unusual subsequence in a time-series. However, as the authors put it, this type of methods lack a reference or normality state and hence do not accurately report outliers.

@@

Frequency-based univariate subsequence outlier detection methods also use a reference period and compute the frequency of each subsequence in the reference or normality period. Hence, a subsequence will be an outlier if it does not appear as frequently as expected in subsequent time periods.

@@

Sampaio et al. present SAMM \cite{SAMM}, an automatic model monitoring system for data streams. Similarly to the method presented in Section \ref{kifer}, SAMM uses two windows: the reference window R and the target window T. Window T contains the last $n_t$ elemnts and window R the elements of a reference period, which is prior to T. SAMM has two main components that interest us: the one responsible for computing the signal and the one that defines the signal threshold.

The Machine Learning (ML) model being monitored outputs a series of scores for each transaction. The authors build two histograms: one with the model scores for the reference window R and another for the target window T. The signal is a measure of similarity between both histograms $H_R$ and $H_T$. The authors considered using the Kolmogorov-Smirnov, Kuiper and Anderson-Darling test but ultimately used the Jensen-Shannon Divergence (JSD) \cite{JSD} to measure similarity between both histograms. The JSD outputs a value between 0 and 1, with totally different histograms being assigned a value of 1. The signal is then computed by applying the JSD metric to both histograms $H_R$ and $H_T$ at each new event.

To compute the threshold for the JSD signal, the authors maintain an histogram of observed JSD values during streaming. To efficiently maintain the JSD histogram, they devise an algorithm called SPEAR which stands for Streaming Percentiles EstimAtoR. SPEAR is linear in time and space complexity regarding the number of histogram bins, which is a small and fixed constant. This way, for a new JSD value measured between the histogram $H_R$ and histogram $H_T$, they compare it with the histogram and alert if above a certain percentile (\textit{e.g.}, the 99th percentile).

@@

Maintaining the target window, \textit{i.e.}, evicting old events and inserting new ones, updating the target histogram $H_T$, computing the JSD between $H_T$ and $H_R$ and checking if it is above a given percentile are all done in constant time. Updating the JSD histogram and dinamically resizing the bins with the SPEAR algorithm takes $O(n)$ time with $n$ as the number of bins. Because the number of bins is small and constant, the update of the JSD histogram is also done in constant time.

@@

\subsubsection*{Applicability to our Hypothesis}
SAMM was designed to work for small time periods. Hence, for the authors, keeping the reference and target windows in memory was no issue. However, in our solution, we need sublinear time and space complexities to handle large windows of data. This implies we can not directly use SAMM to solve our problem.

@@

Even though SAMM is not directly applicable in our context, it provides us with valid and promissing ideas to test.

@@

The De-Amortized Banker’s Aggregator (DABA) \cite{Tangwongsan-DABA} is a SWAG algorithm that allows the computation of \textit{associative} aggregations. This method carefully handles expensive operations and spreads them along the execution. As a resut, the authors are able to turn the average-case behavior into the worst-case behavior. This is commonly known as \textit{de-amortization}, hence the name. The main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout run time.

@@

\subsection*{Time complexity analysis}
Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion has \textit{O(k)} time complexity because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1. Similarly, query answering has \textit{O(k)} time complexity because it needs only to run the input through the \textit{k} hash functions and check if all bits are set to 1. If so, it returns \textit{possibly in set}. Otherwise it returns \textit{definitely not in set}.

@@

% FOR EACH SKETCH DO:
% approx aggregation to compute
% data structures
% algorithm for insertion
% algorithm for query
% error bound related with parameters
% time complexity analysis
% space complexity analysis
% Applicability to our Hypothesis
\subsection*{HyperLogLog}
A couple of years later, the same authors propose HyperLogLog (HLL) \cite{Flajolet-HLL}, an approximate aggregator to compute the cardinality of a set.

@@

The authors claim that the proposed sliding version remains as accurate as of the original HLL aggregator. With an additional memory of only 35kB it is possible to estimate the number of distinct elements in the set with a standard error of 3\% in the presence of up to a million distinct elements.

@@

Despite the fact memory consumption dynamically changes during runtime, this is a valid approach to test our Hypothesis due to the worst-case scenario being a logarithmic space complexity. However, in Section \ref{sec:hll-drv} we analyze an implementation that is not dynamic in memory during run time making it a preferred approach.

@@

In Section \ref{sec:outliers}, we reviewed outlier detection algorithms. However, we found most of them inapplicable in our scenario, because they were resource intensive (both time and memory wise), provided no measure of how divergent and outlier was or because they evolved their reference period.

@@

We then discussed probabilistic data structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}). Despite being efficient, both time and memory wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use to us.

>>>>

The third and final axis is the \textit{outlier type} the method aims to report. The authors identify three distinct outlier types: point outliers, subsequence outliers and time-series outliers. A point outlier corresponds to a single data point in time that deviates from the expected value. These point outliers can be uni or multivariate, depending on the nature of the method. A univariate point outlier is a single time-dependent variable that stands out from the rest of the time-series. On the other hand, a multivariate point outlier takes into account multiple time-dependent variables in comparison to the remaining time-series. Subsequence outliers refer to a sequence of points in time that together show anomalous behavior, but may not be individually considered outliers. Similarly to point outliers, a subsequence outlier may take into account a single time-dependent variable or multiple ones. The authors classify them as univariate subsequence outliers and multivariate subsequence outliers, respectively. Lastly, time-series outliers refer to entire time-series being outliers themselves. This type of outliers can only be detected if the input data is multivariate, to be able to compare multiple time-series.

@@

@@

\subsubsection*{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable to our goals. We need a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, leaving the user confused about why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

@@

Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series to find out the most unusual ones. Each subsequence is compared to the others using a distance metric, for instance, the Euclidean distance. The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in quadratic time complexity. These methods find the most unusual subsequence in a time-series. However, as the authors put it, these methods lack a reference or normality state and hence do not accurately report outliers.

@@

Frequency-based univariate subsequence outlier detection methods also use a reference period and compute the frequency of each subsequence in the reference or normality period. Hence, a subsequence will be an outlier if it does not appear as frequently as expected in subsequent periods.

@@

Sampaio et al. present SAMM \cite{SAMM}, an automatic model monitoring system for data streams. Similarly to the method presented in Section \ref{kifer}, SAMM uses two windows: the reference window R and the target window T. Window T contains the last $n_t$ elements and window R the $n_r$ elements of a reference period, which is prior to T. SAMM has two main components that interest us: the one responsible for computing the signal and the one that defines the signal threshold.

The Machine Learning (ML) model being monitored outputs a series of scores for each transaction. The authors build two histograms: one with the model scores for the reference window R and another for the target window T. The signal is a measure of similarity between both histograms $H_R$ and $H_T$. The authors considered using the Kolmogorov-Smirnov \cite{KolomogorovSmirnov, Kolmogorov-Smirnov}, Kuiper \cite{KuiperTest} and Anderson-Darling \cite{AndersonDarling} test but ultimately used the Jensen-Shannon Divergence (JSD) \cite{JSD} to measure the similarity between both histograms. The JSD outputs a value between 0 and 1, with totally different histograms being assigned a value of 1. The signal is then computed by applying the JSD metric to both histograms $H_R$ and $H_T$ at each new event.

To compute the threshold for the JSD signal, the authors maintain a histogram of observed JSD values during streaming. To efficiently maintain the JSD histogram, they devise an algorithm called SPEAR which stands for Streaming Percentiles EstimAtoR. SPEAR is linear in time and space complexity regarding the number of histogram bins, which is a small and fixed constant. This way, for a new JSD value measured between the histogram $H_R$ and histogram $H_T$, they compare it with the histogram and alert if above a certain percentile (\textit{e.g.}, the 99th percentile).

@@

Maintaining the target window, \textit{i.e.}, evicting old events and inserting new ones, updating the target histogram $H_T$, computing the JSD between $H_T$ and $H_R$ and checking if it is above a given percentile are all done in constant time. Updating the JSD histogram and dynamically resizing the bins with the SPEAR algorithm takes $O(n)$ time with $n$ as the number of bins. Because the number of bins is small and constant, the update of the JSD histogram is also done in constant time.

@@

\subsubsection*{Applicability to our Hypothesis}
SAMM was designed to work for small-time periods. Hence, for the authors, keeping the reference and target windows in memory was no issue. However, in our solution, we need sublinear time and space complexities to handle large windows of data. This implies we can not directly use SAMM to solve our problem.

@@

Even though SAMM is not directly applicable in our context, it provides us with valid and promising ideas to test.

@@

The De-Amortized Banker’s Aggregator (DABA) \cite{Tangwongsan-DABA} is a SWAG algorithm that allows the computation of \textit{associative} aggregations. This method carefully handles expensive operations and spreads them along the execution. As a result, the authors are able to turn the average-case behavior into worst-case behavior. This is commonly known as \textit{de-amortization}, hence the name. The main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout runtime.

@@

\subsection*{Time complexity analysis}
Considering a Bloom Filter with \textit{m} bits and \textit{k} hash functions, insertion has \textit{O(k)} time complexity because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1. Similarly, query answering has \textit{O(k)} time complexity because it needs only to run the input through the \textit{k} hash functions and check if all bits are set to 1. If so, it returns \textit{possibly in set}. Otherwise it returns \textit{definitely not in set}.

@@

\subsection*{HyperLogLog}
A couple of years later, the same authors propose HyperLogLog (HLL) \cite{Flajolet-HLL}, an approximate aggregator to compute the cardinality of a set.

@@

The authors claim that the proposed sliding version remains as accurate as of the original HLL aggregator. With an additional memory of only 35kB, it is possible to estimate the number of distinct elements in the set with a standard error of 3\% in the presence of up to a million distinct elements.

@@

Despite the fact memory consumption dynamically changes during runtime, this is a valid approach to test our Hypothesis due to the worst-case scenario being a logarithmic space complexity. However, in Section \ref{sec:hll-drv} we analyze an implementation that is not dynamic in memory during runtime making it a preferred approach.

@@

In Section \ref{sec:outliers}, we reviewed outlier detection algorithms. However, we found most of them inapplicable in our scenario, because they were resource-intensive (both time and memory-wise), provided no measure of how divergent and outlier was or because they evolved their reference period.

@@

We then discussed probabilistic data structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}). Despite being efficient, both time and memory-wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use to us.

In the next chapter, we formally define the problem we aim to solve and the issues with current state of the art solutions. Then, we formulate our proposal along with the assumptions we make when approaching this problem. Finally, we pose some research questions we aim to answer by the end of this thesis.

### 2020-07-04 chapters/validation.tex
Small corrections

\section{Experiments with Synthetic Data}
In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data. In this section we present summary statistics for the synthetic datasets created and our generated alarms.

\subsection{Single-Feature Analysis}
As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction. In this section we validate our results for single-feature synthetic datasets.

@@

We began by performing single-feature analysis on synthetic data. The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported. Still working with synthetic data, we performed our first multi-feature analysis. In this scenario, we applied the Holm-Bonferroni multiple test correction to the p-values of each hypothesis. Once again, each feature was correctly reported as divergent at the timestamps where we changed the underlying generating distribution.

Finally, we moved on to real data. Due to time constraints and the scope of this thesis, we experimented only with data from one merchant. In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms. A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift. In our last experiment, we attempted to reduce concept drift to a maximum by using the same period as reference and target. To our surprise, we still produced a large number of long lasting alarms. We defer the investigation of the root cause of the problem to future work and leave a few untested hypothesis as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data. In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset. Additionally, our sampling may have been insufficient. To test this, an experiment with lower $\gamma$ and hence higher number of samples should be carried out.

>>>>

\section{Experiments with Synthetic Data}
In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data. In this section, we present summary statistics for the synthetic datasets created and our generated alarms.

\subsection{Single-Feature Analysis}
As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction. In this section, we validate our results for single-feature synthetic datasets.

@@

We began by performing a single-feature analysis of synthetic data. The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported. Still working with synthetic data, we performed our first multi-feature analysis. In this scenario, we applied the Holm-Bonferroni multiple test correction to the p-values of each hypothesis. Once again, each feature was correctly reported as divergent at the timestamps where we changed the underlying generating distribution.

Finally, we moved on to real data. Due to time constraints and the scope of this thesis, we experimented only with data from one merchant. In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms. A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift. In our last experiment, we attempted to reduce concept drift to a maximum by using the same period as reference and target. To our surprise, we still produced a large number of long-lasting alarms. We defer the investigation of the root cause of the problem to future work and leave a few untested hypotheses as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data. In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset. Additionally, our sampling may have been insufficient. To test this, an experiment with a lower $\gamma$ and hence a higher number of samples should be carried out.
